quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-05-12 paper-quollr.Rmd
Start working on the application

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], create wildly different representations depending on the selected method and (hyper)parameter choices (Figure \@ref(fig:example)). It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

```{r fig-example, fig.alt = "NLDR representations with different methods and (hyper)parameter choices.", fig.cap = "Six different NLDR representations of Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles have been generated using different techniques and different (hyper)parameter choices. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical (hyper)parameter choice. Would their downstream analysis decisions vary based on the version they observe? Which representation most accurately represents the structure in high-dimensional space?", out.width = "100%", fig.width= 6, fig.height= 4}

@@

Need to find another application

@@

training_data_limb <- read_rds("data/limb_muscles/facs_limb_muscles_pcs_10.rds")

@@

```{r}
#| label: tsne-num-bins-limb
#| eval: false

## Compute hexbin parameters
num_bins_x_limb <- 19

algo_obj_limb <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_limb, 
  nldr_data = tsne_limb, 
  num_x_bins = num_bins_x_limb)

tsne_limb_scaled <- algo_obj_limb$nldr_scaled
distance_limb <- algo_obj_limb$distance_df
tr_from_to_df_limb <- algo_obj_limb$tr_from_to_df
benchmark_limb <- algo_obj_limb$benchmark
df_bin_centroids_limb <- algo_obj_limb$df_bin_centroids
df_bin_limb <- algo_obj_limb$df_bin

distance_df_small_edges_limb <- distance_limb |>
  filter(distance < 0.1) #benchmark_limb

tr_from_to_df_limb <- right_join(
  tr_from_to_df_limb, distance_df_small_edges_limb,
  by = c("from", "to"))

tsne_limb_scaled_with_cluster <- inner_join(tsne_limb_scaled, cluster_df, by = "ID") |>
  mutate(cluster.ids = as.character(cluster.ids))

trimesh_removed_limb <- ggplot() + 
  geom_point(
    data = tsne_limb_scaled_with_cluster,
    aes(
      x = emb1,
      y = emb2,
      color = cluster.ids
    ),
    alpha = 0.3#,
    #color = "#a65628"
  ) +
    geom_segment(data = tr_from_to_df_limb, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494')) +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: prep-limb-tsne-author-model-proj
#| eval: false

data_limb <- training_data_limb |> 
  select(-ID) |>
  mutate(type = "data")

df_b_limb <- df_bin_limb |>
  dplyr::filter(hb_id %in% df_bin_centroids_limb$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_limb <- df_b_limb[match(df_bin_centroids_limb$hexID, df_b_limb$hb_id),] |>
  dplyr::select(-hb_id)

# Apply the scaling
df_model_data_limb <- bind_rows(data_limb, df_b_limb)
scaled_limb <- scale_data_manual(df_model_data_limb, "type") |>
  as_tibble()

scaled_limb_data <- scaled_limb |>
  filter(type == "data") |>
  select(-type)

scaled_limb_data_model <- scaled_limb |>
  filter(type == "model") |>
  select(-type)
```

```{r}
#| label: langevitour-limb-tsne-author-proj
#| eval: false

data_limb_n <- data_limb |>
  select(-type) |>
  mutate(type = as.character(tsne_limb_scaled_with_cluster$cluster.ids))

df_model_data_limb_n <- bind_rows(df_b_limb, data_limb_n)

langevitour::langevitour(df_model_data_limb_n[1:(length(df_model_data_limb_n)-1)],
                         lineFrom = distance_df_small_edges_limb$from,
                         lineTo = distance_df_small_edges_limb$to,
                         group = factor(df_model_data_limb_n$type,
                                        c("0", "1", "2", "3", "4", "5", "6", "model")),
                         levelColors = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494', "#000000"))
```

```{r}
#| label: tsne-best-num-bins-limb
#| eval: false

## Compute hexbin parameters
num_bins_x_limb <- 18

algo_obj_limb <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_limb, 
  nldr_data = tsne_limb2, 
  num_x_bins = num_bins_x_limb)

tsne_limb_scaled_best <- algo_obj_limb$nldr_scaled
distance_limb <- algo_obj_limb$distance_df
tr_from_to_df_limb <- algo_obj_limb$tr_from_to_df
benchmark_limb <- algo_obj_limb$benchmark
df_bin_centroids_limb <- algo_obj_limb$df_bin_centroids
df_bin_limb <- algo_obj_limb$df_bin

distance_df_small_edges_limb <- distance_limb |>
  filter(distance < benchmark_limb) #benchmark_limb

tr_from_to_df_limb <- right_join(
  tr_from_to_df_limb, distance_df_small_edges_limb,
  by = c("from", "to"))

tsne_limb_scaled_best_with_cluster <- inner_join(tsne_limb_scaled_best, cluster_df, by = "ID") |>
  mutate(cluster.ids = as.character(cluster.ids))

trimesh_removed_limb_best <- ggplot() + 
  geom_point(
    data = tsne_limb_scaled_best_with_cluster,
    aes(
      x = emb1,
      y = emb2,
      color = cluster.ids
    ),
    alpha = 0.3#,
    #color = "#ff7f00"
  ) +
    geom_segment(data = tr_from_to_df_limb, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494')) +
  interior_annotation("b1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: prep-limb-tsne-best-model-proj
#| eval: false

data_limb <- training_data_limb |> 
  select(-ID) |>
  mutate(type = "data")

df_b_limb <- df_bin_limb |>
  dplyr::filter(hb_id %in% df_bin_centroids_limb$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_limb <- df_b_limb[match(df_bin_centroids_limb$hexID, df_b_limb$hb_id),] |>
  dplyr::select(-hb_id)

# Apply the scaling
df_model_data_limb <- bind_rows(data_limb, df_b_limb)
scaled_limb <- scale_data_manual(df_model_data_limb, "type") |>
  as_tibble()

scaled_limb_data <- scaled_limb |>
  filter(type == "data") |>
  select(-type)

scaled_limb_data_model <- scaled_limb |>
  filter(type == "model") |>
  select(-type)
```

```{r}
#| label: langevitour-limb-tsne-best-proj
#| eval: false

df_model_data_limb_n <- bind_rows(df_b_limb, data_limb_n)

langevitour::langevitour(df_model_data_limb_n[1:(length(df_model_data_limb_n)-1)],
                         lineFrom = distance_df_small_edges_limb$from,
                         lineTo = distance_df_small_edges_limb$to,
                         group = factor(df_model_data_limb_n$type,
                                        c("0", "1", "2", "3", "4", "5", "6", "model")),
                         levelColors = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854','#ffd92f','#e5c494', "#000000"))
```

```{r}
#| eval: false
trimesh_removed_limb + trimesh_removed_limb_best +
  plot_layout(ncol = 3)
```

>>>>

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], create wildly different representations depending on the selected method and (hyper)parameter choices (Figure @fig-nldrlayouts). It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

```{r fig-nldrlayouts, fig.alt = "NLDR representations with different methods and (hyper)parameter choices.", fig.cap = "Six different NLDR representations of Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles have been generated using different techniques and different (hyper)parameter choices. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical (hyper)parameter choice. Would their downstream analysis decisions vary based on the version they observe? Which representation most accurately represents the structure in high-dimensional space?", out.width = "100%", fig.width= 6, fig.height= 4}

@@

Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the discovered clusters, and help to understand the results. The purpose of this example is to *illustrate how to use our method to help decide on an appropriate NLDR layout that accurately represents the data*.

The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better \gD{} layout, as needed.

@@

data_limb <- read_rds("data/limb_muscles/facs_limb_muscles_pcs_10.rds")

@@

```{r}
#| label: tsne-model-limb

### 2025-05-12 paper-quollr.tex
Start working on the application

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}, create wildly different representations depending on the selected method and (hyper)parameter choices (Figure \ref{fig:example}). It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

\begin{figure}
\includegraphics[width=1\linewidth,alt={NLDR representations with different methods and (hyper)parameter choices.}]{paper-quollr_files/figure-latex/fig-example-1} \end{figure}

>>>>

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}, create wildly different representations depending on the selected method and (hyper)parameter choices (Figure \citet{fig-nldrlayouts}). It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

\begin{figure}
\includegraphics[width=1\linewidth,alt={NLDR representations with different methods and (hyper)parameter choices.}]{paper-quollr_files/figure-latex/fig-nldrlayouts-1} \caption{Six different NLDR representations of Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles have been generated using different techniques and different (hyper)parameter choices. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical (hyper)parameter choice. Would their downstream analysis decisions vary based on the version they observe? Which representation most accurately represents the structure in high-dimensional space?}\label{fig:fig-nldrlayouts}
\end{figure}
